Demes: a standard format for demographic models

### 2022-05-27 paper.tex
Americanize the English.

\begin{abstract}
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being proposed and tested.
Demographic models of current interest
typically consist of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs, which can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labour intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.
\end{abstract}

@@

Simulation is a core tool in population genetics, and
many methods have been developed over the past three
decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models,
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.,][]{hudson2002generating,hernandez_flexible_2008,kern2016discoal},
a custom input file
format~\citep[e.g.,][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models to be defined programmatically~\citep[e.g.,][]{
thornton2014cpp,hernandez_sfs_code_2015,kelleher2016efficient,
becheler2019quetzal,haller2019slim,thornton2019-nu,baumdicker2021-iu}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors likely.
Input parameter file formats for simulators allow the model specification
to be less terse and allow for documentation in the form of comments.
Several Graphical User Interface (GUI) and visualisation
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
parreira_spams_2009,ewing2010msms,parobek_skelesim_2017,zhou2018popdemog}.
However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the  syntax of a specific simulator.
There is currently no way in which demographic models inferred by
different packages can be simulated or visualised by downstream software.

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab:software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little programming effort.

@@

\CatchFileDef{\softwaretable}{software-table.tex}{}
\renewcommand{\arraystretch}{1.5}
\begin{table}
    \begin{center}
        \begin{tabular}{lp{12cm}}
            \softwaretable
        \end{tabular}
    \end{center}
    \caption{
        \label{tab:software}
        \textbf{Software support for Demes.}
        We have included software infrastructure developed
        for working with Demes models (such as parsing, validation,
        and visualisation) as well as downstream
        software that implement the specification,
        at the time of writing.
    }
\end{table}

@@

The Demes specification is static by design---we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.
It is important to note, however, that Demes may be used in applications that
include additional population genetic processes outside of what is explicitly
modelled in the specification, such as interpreting population sizes as
carrying capacities, implementations of hard selection, or layering more
complicated mating or spatial structure. The Demes specification therefore
provides a basic model that can be elaborated on where necessary.

@@

\section*{Discussion}
Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been ad-hoc, and conversions between the many different ways
of expressing such models is both labour intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab:software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that consume, transform, or produce this format.

Reproducibility is a significant problem throughout the
sciences~\citep{baker20161}, and various measures have been
proposed to increase the likelihood of researchers being
able to replicate results in the
literature~\citep{munafo2017manifesto}. The most basic requirement
for reproducibility is that we must be able to state precisely what
the result in question is. The lack of standardisation in how
complex demographic models are communicated today, and the lack of
precision in the published model descriptions means that it is difficult
to replicate analyses, or reproduce those models for later simulation.
Thus, we hope that the Demes standard introduced here will be widely adopted
by simulation and inference methods and be used for reporting results in
publications, either as supplemental material or uploaded to a data repository.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and formally defined using
JSON Schema~\citep{wright2020json}.
Along with the schema, full technical details of the
of the model are provided in the
online specification document
(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
This specification rigorously defines the data model,
fully describing the entities and their
relationships, and the required behaviour of implementations.
Since the online specification is definitive,
we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

Below we detail the population genetics model in
Appendix~\ref{sec:appendix-pop-gen-model},
the high-level, human-readable model specification in
Appendix~\ref{sec:appendix-spec}, demographic model parsers and their
intended behaviour in Appendix~\ref{sec:appendix-parsers}, and the
intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
The extended figures in Appendix~\ref{sec:appendix-figures} provide
additional examples and illustrations of Demes usage.

@@

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', understood to be a collection of individuals that can be
conveniently modelled using a defined set of rules and parameters
\citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either according to their ancestor-descendant relationships
or through processes involving migrations.
Few other properties of the populations are specified in the model:
we are concerned primarily with defining the populations, their sizes, and the
movement of individuals between those populations.

@@

Each population has an assigned selfing rate and cloning rate, where each defines
the probability that offspring are generated from one generation to the next by either
self-fertilisation or cloning of an individual.
More specifically, for a given epoch within a population denote the clonal rate by
$\sigma$ and the selfing rate by $S$.
$S$ and $\sigma$ can take any value between zero and one and
can sum to more than one.
Each generation a proportion of offspring
$\sigma$ are expected to be generated through clonal reproduction,
while $1-\sigma$ are expected to arise through sexual reproduction.
Within the sexually-reproduced offspring,
$S$ are born via self-fertilisation while the rest
have parents drawn at random from the previous generation.
Depending on the simulator, this random drawing of parent may occur
either with or without replacement. When drawing occurs with replacement, a small
amount of ``residual'' selfing is expected, so that the realized selfing probability
is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
and $S=0$, selfing may still occur with probability $1/N$), although
this effect is negligible in large populations \citep{nordborg1997coalescent}.
Simulators that allow variable rates of selfing are expected to clearly
document the expected behaviour.

@@

The Demes Human Data Model (HDM) is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
serialisation language with an emphasis on simplicity and which interoperates well
with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalence with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

@@

The Demes specification precisely defines the required behaviour of parsers,
which translate a model described in the HDM into the MDM, where values have been
assigned to all parameters and data
constraints have been checked. The output is formally defined as JSON, but in
practice parsers output an object model that is suitable for the particular
programming language and that can be used directly by the implementing program.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

Similarly, one might be interested in running simulations in which
the demographic model parameters are drawn from a distribution, e.g.,
in ABC inference~\citep{beaumont2002approximate}.
Other inference procedures based on optimising a loss function
\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}
%(\ggcomment{more refs? MCMC? neural networks?})
need users to specify parameter bounds,
and possibly non-linear or conditional constraints between parameters.
Indeed, the choice of how to parameterise a model could be important for
some inference methods (e.g. absolute times versus relative times between events).

>>>>

\begin{abstract}
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being proposed and tested.
Demographic models of current interest
typically consist of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs, which can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labor intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.
\end{abstract}

@@

Simulation is a core tool in population genetics, and
many methods have been developed over the past three
decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealized population models,
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.,][]{hudson2002generating,hernandez_flexible_2008,kern2016discoal},
a custom input file
format~\citep[e.g.,][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models to be defined programmatically~\citep[e.g.,][]{
thornton2014cpp,hernandez_sfs_code_2015,kelleher2016efficient,
becheler2019quetzal,haller2019slim,thornton2019-nu,baumdicker2021-iu}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors likely.
Input parameter file formats for simulators allow the model specification
to be less terse and allow for documentation in the form of comments.
Several Graphical User Interface (GUI) and visualization
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
parreira_spams_2009,ewing2010msms,parobek_skelesim_2017,zhou2018popdemog}.
However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the  syntax of a specific simulator.
There is currently no way in which demographic models inferred by
different packages can be simulated or visualized by downstream software.

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behavior of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualization of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab:software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little programming effort.

@@

\CatchFileDef{\softwaretable}{software-table.tex}{}
\renewcommand{\arraystretch}{1.5}
\begin{table}
    \begin{center}
        \begin{tabular}{lp{12cm}}
            \softwaretable
        \end{tabular}
    \end{center}
    \caption{
        \label{tab:software}
        \textbf{Software support for Demes.}
        We have included software infrastructure developed
        for working with Demes models (such as parsing, validation,
        and visualization) as well as downstream
        software that implement the specification,
        at the time of writing.
    }
\end{table}

@@

The Demes specification is static by design---we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.
It is important to note, however, that Demes may be used in applications that
include additional population genetic processes outside of what is explicitly
modeled in the specification, such as interpreting population sizes as
carrying capacities, implementations of hard selection, or layering more
complicated mating or spatial structure. The Demes specification therefore
provides a basic model that can be elaborated on where necessary.

@@

\section*{Discussion}
Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been \textit{ad-hoc}, and conversions between the many different ways
of expressing such models is both labor intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab:software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that consume, transform, or produce this format.

Reproducibility is a significant problem throughout the
sciences~\citep{baker20161}, and various measures have been
proposed to increase the likelihood of researchers being
able to replicate results in the
literature~\citep{munafo2017manifesto}. The most basic requirement
for reproducibility is that we must be able to state precisely what
the result in question is. The lack of standardization in how
complex demographic models are communicated today, and the lack of
precision in the published model descriptions means that it is difficult
to replicate analyses, or reproduce those models for later simulation.
Thus, we hope that the Demes standard introduced here will be widely adopted
by simulation and inference methods and be used for reporting results in
publications, either as supplemental material or uploaded to a data repository.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and formally defined using
JSON Schema~\citep{wright2020json}.
Along with the schema, full technical details of the
of the model are provided in the
online specification document
(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
This specification rigorously defines the data model,
fully describing the entities and their
relationships, and the required behavior of implementations.
Since the online specification is definitive,
we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

Below we detail the population genetics model in
Appendix~\ref{sec:appendix-pop-gen-model},
the high-level, human-readable model specification in
Appendix~\ref{sec:appendix-spec}, demographic model parsers and their
intended behavior in Appendix~\ref{sec:appendix-parsers}, and the
intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
The extended figures in Appendix~\ref{sec:appendix-figures} provide
additional examples and illustrations of Demes usage.

@@

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', understood to be a collection of individuals that can be
conveniently modeled using a defined set of rules and parameters
\citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either according to their ancestor-descendant relationships
or through processes involving migrations.
Few other properties of the populations are specified in the model:
we are concerned primarily with defining the populations, their sizes, and the
movement of individuals between those populations.

@@

Each population has an assigned selfing rate and cloning rate, where each defines
the probability that offspring are generated from one generation to the next by either
self-fertilization or cloning of an individual.
More specifically, for a given epoch within a population denote the clonal rate by
$\sigma$ and the selfing rate by $S$.
$S$ and $\sigma$ can take any value between zero and one and
can sum to more than one.
Each generation a proportion of offspring
$\sigma$ are expected to be generated through clonal reproduction,
while $1-\sigma$ are expected to arise through sexual reproduction.
Within the sexually-reproduced offspring,
$S$ are born via self-fertilization while the rest
have parents drawn at random from the previous generation.
Depending on the simulator, this random drawing of parent may occur
either with or without replacement. When drawing occurs with replacement, a small
amount of ``residual'' selfing is expected, so that the realized selfing probability
is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
and $S=0$, selfing may still occur with probability $1/N$), although
this effect is negligible in large populations \citep{nordborg1997coalescent}.
Simulators that allow variable rates of selfing are expected to clearly
document the expected behavior.

@@

The Demes Human Data Model (HDM) is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
serialization language with an emphasis on simplicity and which interoperates well
with JSON (indeed, YAML 1.2 is a super set of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalence with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

@@

The Demes specification precisely defines the required behavior of parsers,
which translate a model described in the HDM into the MDM, where values have been
assigned to all parameters and data
constraints have been checked. The output is formally defined as JSON, but in
practice parsers output an object model that is suitable for the particular
programming language and that can be used directly by the implementing program.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (published under liberal open source licenses),
providing a solid foundation for the software ecosystem.

@@

Similarly, one might be interested in running simulations in which
the demographic model parameters are drawn from a distribution, e.g.,
in ABC inference~\citep{beaumont2002approximate}.
Other inference procedures based on optimising a loss function
\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}
%(\ggcomment{more refs? MCMC? neural networks?})
need users to specify parameter bounds,
and possibly non-linear or conditional constraints between parameters.
Indeed, the choice of how to parameterize a model could be important for
some inference methods (e.g. absolute times versus relative times between events).
